Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-11-11 fstats_pca_lewontin.tex
supp files, updated fig

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

@@

>>>>

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that just using the first few PCs provides good population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

@@

\section{Supplementary File Description}
Both supplementary files give the individual-identifier, sex and population label for each sample.
